---
title: "Untitled"
output: html_document
date: "2024-10-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(actuar)
library(fitdistrplus)
library(dplyr)
library(distr)
library(DescTools)
library(bp)
library(car)
```

# The reparameterized inverse Gaussian distribution{.tabset}

## Varying $\theta$

```{r}
# Clean Component 
clnrIG    <- function(x, t, g, a, log = F){ (a)*dinvgauss(x,sqrt(t*(3*g+t)), (t*(3*g+t))/g,log=log)}

# contaminated component
contrIG    <- function(x, t, g, a, e, log = F){ (1-a)*dinvgauss(x,sqrt(t*(3*g*e+t)), (t*(3*g*e+t))/(g*e),log=log)}

curve(clnrIG(x, 0.4, 1, a = 1), 0, 12, ylim = c(0,1), ylab = "f(x)", xlab = expression(x), lty = 2 )
curve(clnrIG(x, 1, 1, a = 1), 0, 12, add = TRUE)
curve(clnrIG(x, 4, 1, a = 1), 0, 12, add = TRUE, lty = 3)
legend("topright", legend =c(expression(theta~"= 0.4"), expression(theta~"= 1"), expression(theta~"= 4")), lty = c(2, 1, 3), title = expression("rIG with varying"~ theta))
```
## Varying $\gamma$

```{r}

curve(clnrIG(x, 1, 0.4, a = 1), 0, 15, ylim = c(0,0.8), ylab = "f(x)", xlab = expression(x), lty = 2)
curve(clnrIG(x, 1, 1, a = 1), 0, 15, add = TRUE)
curve(clnrIG(x, 1, 4, a = 1), 0, 15, add = TRUE, lty = 3)
legend("topright", legend =c(expression(gamma~"= 0.4"), expression(gamma~"= 1"), expression(gamma~"= 4")), lty = c(2, 1, 3), title = expression("rIG with varying"~ gamma)) 

```
# Simulation from base rIG and classical MLE

```{r}

# mrIG

dminvgauss <- function(x,theta,v,log=F){
dinvgauss(x,sqrt(theta*(3*v+theta)), (theta*(3*v+theta))/v,log=log)
}



theta <- 2
v <- 0.4
#simulate data

x<-rinvgauss(10000,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v)

# log-like with constrained parameters 
dminvgaussMLE <- function(par,x){
theta <- exp(par[1]) #back-transformation
v <- exp(par[2]) #back-transformation
return(sum(dminvgauss(x, theta, v, log = T)))
}

start <- log(c(theta,v))

est <- optim(par = start,fn = dminvgaussMLE,x=x, control=list(fnscale= -1))

par <- est$par
thetaest <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

curve(clnrIG(x, theta, v, a = 1), 0, 15, ylim = c(0,0.5), ylab = "f(x)", xlab = expression(x))
curve(clnrIG(x, thetaest, vest, a = 1), add = TRUE, lty = 2)
legend("topright", legend =c("Empirical rIG density", "ML Estimated density"), lty = c(1,2), title = "Simulated & Estimated rIG") 


```
# Mixed Simulation and base rIG estimation{.tabset}

## $\theta = 2$, $\gamma = 0.4$, $\alpha = 0.75$ and $\eta = 2$

```{r}

theta <- 2
v <- 0.4
vinf <- 0.4*2 # inflate the variance by multiplying by eta

set.seed(18)
x<- c(rinvgauss(10000*0.75,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v), rinvgauss(10000*0.25,sqrt(theta*(3*vinf+theta)),(theta*(3*vinf+theta))/vinf))


# mrIG

dminvgauss <- function(x,theta,v,log=F){
dinvgauss(x,sqrt(theta*(3*v+theta)), (theta*(3*v+theta))/v,log=log)
}


dminvgaussMLE <- function(par,x){
theta <- exp(par[1]) #back-transformation
v <- exp(par[2]) #back-transformation
return(sum(dminvgauss(x, theta, v, log = T)))
}

start <- log(c(theta,v))

est <- optim(par = start,fn = dminvgaussMLE,x=x, control=list(fnscale= -1))

par <- est$par
thetaest <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

curve(clnrIG(x, 2, 0.4, a = 0.75) + contrIG(x, 2, 0.4, a = 0.75, 2), 0, 15, ylab = "f(x)", ylim = c(0, 0.5), xlab = expression(x))
curve(clnrIG(x, thetaest, vest, a = 1), add = TRUE, lty = 2)
legend("topright", legend =c("Mixed rIG density", "ML Estimated density"), lty = c(1,2), title = "Simulated & Estimated rIG") 

```
## $\theta = 2$, $\gamma = 0.4$, $\alpha = 0.85$ and $\eta = 4$

```{r}

theta <- 2
v <- 0.4
vinf <- 0.4*4

set.seed(18)
x<- c(rinvgauss(10000*0.85,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v), rinvgauss(10000*0.15,sqrt(theta*(3*vinf+theta)),(theta*(3*vinf+theta))/vinf))

start <- log(c(theta,v))

est <- optim(par = start,fn = dminvgaussMLE,x=x, control=list(fnscale= -1))

par <- est$par
thetaest <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

curve(clnrIG(x, 2, 0.4, a = 0.85) + contrIG(x, 2, 0.4, a = 0.85, 4), 0, 15, ylab = "f(x)", ylim = c(0, 0.5), xlab = expression(x))
curve(clnrIG(x, thetaest, vest, a = 1), add = TRUE, lty = 2)
legend("topright", legend =c("Mixed rIG density", "ML Estimated density"), lty = c(1,2), title = "Simulated & Estimated rIG") 

```
# Mixed simulation & mixed estimation by ECM 

```{r,  warning = FALSE}

# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}

# mode-parameterized IG
rIG <- function(x,theta,v,log=F){
dinvgauss(x,sqrt(theta*(3*v+theta)), (theta*(3*v+theta))/v,log=log)
}

z <- function(x, t, g, a, e){
  return(ifelse((a*rIG(x, t, g))/(clnrIG(x, t, g, a, log = F) + contrIG(x, t, g, a, e, log = F)) > 0.5, 1, 0))
}

v <- 0.4
theta <- 2
alpha <- 0.75
eta <- 2
vinf <- 2*0.4

set.seed(18)
sim <- c(rinvgauss(1000*0.75,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v), rinvgauss(1000*0.25,sqrt(theta*(3*vinf+theta)),(theta*(3*vinf+theta))/vinf))

good <- z(sim, 2, 0.4, 0.75, 2)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.75, 2, 0.4, 2)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.75
theta_1 <- 2
gamma_1 <- 0.4
eta_1 <- 2


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0000001 & k <= 150) {
  # E step
  good <- z(sim,theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dminvgaussmixecm <- function(par){
  t <- exp(par[1])
  g <- exp(par[2])
  e <- exp(par[3])
  sum(good*rIG(sim, t, g, log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- exp(optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$par[1:3])
  
  search[k, 6] <- optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$value
  
  alpha_1 <- search[k,2]
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  loglike <- search[k, 6]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}

curve(clnrIG(x, 2, 0.4, a = 0.75) + contrIG(x, 2, 0.4, a = 0.75, 2), 0, 20, ylab = "", ylim = c(0, 0.5))
curve(clnrIG(x, theta_1, gamma_1, a = alpha_1), 0, 20, add = TRUE, lty = 2)
#curve(contrIG(x, theta_1, gamma_1, a = alpha_1, eta_1), 0, 20, add = TRUE, lty = 3)
legend("topright", legend =c("Mixed rIG density", "good points density"), lty = c(1,2), title = "Simulated & Estimated rIG") 


```

```{r,  warning = FALSE}

# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}


z <- function(x, t, g, a, e){
  return(ifelse((a*rIG(x, t, g))/(clnrIG(x, t, g, a, log = F) + contrIG(x, t, g, a, e, log = F)) > 0.5, 1, 0))
}

v <- 0.4
theta <- 2
alpha <- 0.85
eta <- 4
vinf <- 4*v

set.seed(18)
sim <- c(rinvgauss(1000*0.85,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v), rinvgauss(1000*0.15,sqrt(theta*(3*vinf+theta)),(theta*(3*vinf+theta))/vinf))

good <- z(sim, 2, 0.4, 0.85, 4)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.85, 2, 0.4, 4)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.85
theta_1 <- 2
gamma_1 <- 0.4
eta_1 <- 4


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0000001 & k <= 150) {
  # E step
  good <- z(sim,theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dminvgaussmixecm <- function(par){
  t <- exp(par[1])
  g <- exp(par[2])
  e <- exp(par[3])
  sum(good*rIG(sim, t, g, log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- exp(optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$par[1:3])
   search[k, 6] <- optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$value
  
  alpha_1 <- search[k,2]
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  loglike <- search[k, 6]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}

curve(clnrIG(x, 2, 0.4, a = 0.85) + contrIG(x, 2, 0.4, a = 0.85, 4), 0, 12, ylab = "f(x)", ylim = c(0, 0.5), xlab = "Simulated Mixture")
curve(clnrIG(x, theta_1, gamma_1, a = alpha_1), 0, 50, add = TRUE, lty = 2)
curve(contrIG(x, theta_1, gamma_1, a = alpha_1, eta_1), 0, 50, add = TRUE, lty = 3)
legend("topright", legend =c("Mixed rIG density", "good points density", "bad points density"), lty = c(1,2, 3), title = "Simulated & Estimated rIG") 


```

# Application, Real Data Analysis of Sleep Blood Pressure data{.tabset} 

## Exploratory Data Analysis


```{r,  warning = FALSE}



# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}

z <- function(x, t, g, a, e){
  return(ifelse((a*rIG(x, t, g))/(clnrIG(x, t, g, a, log = F) + contrIG(x, t, g, a, e, log = F)) > 0.7, 1, 0))
}


v <- 130
theta <- 66
alpha <- 0.97
eta <- 110 # just guessing 
vinf <- v*110

set.seed(18)
sim <- c(rinvgauss(26387*alpha,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v), rinvgauss(26387*(1-alpha),sqrt(theta*(3*vinf+theta)),(theta*(3*vinf+theta))/vinf))

good <- z(sim, theta, v, 0.97, 110)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.97, theta, v, 110)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.97
theta_1 <- 66
gamma_1 <- 130
eta_1 <-110


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0000001 & k <= 150) {
  # E step
  good <- z(sim,theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dminvgaussmixecm <- function(par){
  t <- exp(par[1])
  g <- exp(par[2])
  e <- exp(par[3])
  sum(good*rIG(sim, t, g, log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- exp(optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$par[1:3])
   search[k, 6] <- optim(log(c(theta_1, gamma_1, eta_1)), dminvgaussmixecm, control=list(fnscale= -1))$value
  
  alpha_1 <- round(search[k,2], 5)
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  loglike <- search[k, 6]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}



BIC <- -2*min(Q$X2) + 4*26386
AIC <- 2*4 - 2*min(Q$X2)

```

```{r}



hist(income$Amount_invested_monthly, freq = FALSE, breaks = 30, xlab = "Amount Invested Monthly (in USD)", main = "", ylim = c(0,0.006), col = "white")
lines(density(income$Amount_invested_monthly))
rug(income$Amount_invested_monthly,side=1,col=income$Good)
#curve(clnrIG(x, 2, 0.4, a = 0.85) + contrIG(x, 2, 0.4, a = 0.85, 4), 0, 50, ylab = "", ylim = c(0, 0.1))
curve(clnrIG(x, theta_1, gamma_1, a = alpha_1), 0, 2000, lty = 2, add = TRUE)
curve(contrIG(x, theta_1, gamma_1, a = alpha_1, eta_1), 0, 2000, add = TRUE, lty = 3)
legend("topright", legend =c("Actual data density", "good points density", "bad points density"), lty = c(1,2, 3), title = "Simulated & Estimated rIG") 

```
### Posterior probability to be good 

```{r}

# mode-parameterized IG
rIG <- function(x,theta,v,log=F){
dinvgauss(x,sqrt(theta*(3*v+theta)), (theta*(3*v+theta))/v,log=log)
}

# Clean Component 
clnrIG    <- function(x, t, g, a, log = F){ (a)*dinvgauss(x,sqrt(t*(3*g+t)), (t*(3*g+t))/g,log=log)}

# contaminated component
contrIG    <- function(x, t, g, a, e, log = F){ (1-a)*dinvgauss(x,sqrt(t*(3*g*e+t)), (t*(3*g*e+t))/(g*e),log=log)}

sort <- sort(income$Amount_invested_monthly)

post <- (alpha_1*rIG(sort, theta_1, gamma_1))/(clnrIG(sort, theta_1, gamma_1, alpha_1, log = F) + contrIG(sort, theta_1, gamma_1, alpha_1, eta_1, log = F))

good <- ifelse(post > 0.5, 1, 0)

plot(sort, post, pch = 16, col = good)

income <- na.omit(income)

plot(income$Amount_invested_monthly, income$Posterior, ylim = c(0,1), main = "", xlab = "Amount Invested Monthly (in USD)", ylab = "posterior probability to be good", col = income$Good)

income2 <- tibble(bind_cols(sort,post,good))

income2$...3 <- as.integer(income2$...3)

ggplot(income2,aes(x=...1,y=...2,col=as.factor(...3)))+geom_point() + theme_bw() +
  theme(axis.line = element_line(colour = "black"),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    panel.border = element_rect(colour = "black", size = 1),
    panel.background = element_blank() )+ 
    theme(legend.position="none") + 
   xlab("Amount Invested Monthly (in USD)") +
   ylab("posterior probability to be good") +
geom_hline(yintercept = 0.5, linetype = "dotted") + 
  geom_hline(yintercept = 1.0) + 
  geom_hline(yintercept = 0.0) + 
  scale_y_continuous(breaks = seq(0, 1, by = 0.1))+ 
  scale_x_continuous(breaks = seq(0, 2000, by =250))+ 
  scale_color_manual(values = c("black", "grey"))


```


## Fit base

```{r}

theta <- 66
v <- 130

set.seed(18)
x<- c(rinvgauss(26386,sqrt(theta*(3*v+theta)),(theta*(3*v+theta))/v))

start <- log(c(theta,v))

est <- optim(par = start,fn = dminvgaussMLE,x=x, control=list(fnscale= -1))

par <- est$par
thetaest <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

hist(income$Amount_invested_monthly, freq = FALSE, breaks = 30, xlab = "Amount Invested Monthly (in USD)", main = "", ylim = c(0,0.007), col = "white")
lines(density(income$Amount_invested_monthly))
rug(income$Amount_invested_monthly,side=1,col="black")
#curve(clnrIG(x, 2, 0.4, a = 0.85) + contrIG(x, 2, 0.4, a = 0.85, 4), 0, 15, ylab = "f(x)", ylim = c(0, 0.5), xlab = expression(x))
curve(clnrIG(x, thetaest, vest, a = 1), add = TRUE, lty = 2)
legend("topright", legend =c("Actual Data density", "ML Estimated density"), lty = c(1,2), title = "Simulated & Estimated rIG")

BIC <- -2*loglike + 4*26386
AIC <- 2*2 - 2*loglike
```



































































































































































```{r Inputation & EDA}

data("bp_hypnos")

## Assign the output of the process_data function to a new dataframe object
hypnos_proc <- process_data(bp_hypnos,
                            bp_type = "abpm",
                            sbp = 'syst',
                            dbp = 'diast',
                            date_time = 'date.time',
                            id = 'id',
                            visit = 'visit',
                            hr = 'hr',
                            wake = 'wake',
                            pp = 'pp',
                            map = 'map',
                            rpp = 'rpp')

# mrIG

dmgamma <- function(x,theta,v,log=F){
dgamma(x,theta/v+1,v,log=log)
}

# input data 
set.seed(21)
x <- hypnos_proc$SBP

theta < as.numeric(Mode(x))
v <- sd(x)**2

# log-like with constrained parameters 
dmgammaMLE <- function(par,x){
theta <- exp(par[1]) #back-transformation
v <- exp(par[2]) #back-transformation
return(sum(dmgamma(x,theta,v,log=T)))
}

start <- log(c(theta,v))

est <- optim(par = start,fn = dmgammaMLE,x=x, control=list(fnscale= -1))

par <- est$par
thetaest <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

hist(x, main = "", xlim = c(75, 200), freq = FALSE, breaks = 20)
lines(density(x))
rug(x,side=1,col="black")
curve(clnrIG(x, thetaest, vest, a = 1), 60, 200, ylim = c(0,0.025), ylab = "f(x)", xlab = expression(x), add = TRUE, lty = 2)
legend("topright", legend =c("Actual data density", "ML Estimated density"), lty = c(1,2), title = "Actual data & Estimated rIG") 


```

```{r,  warning = FALSE}

# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}

# reparameterised rIG pdf 
rIG <- function(x, t, g, log = F){dgamma(x,t/g+1,g,log=log)}

#clean component
clnrIG    <- function(x, t, g, a, log = F){ (a)*dgamma(x,t/g+1,g,log=log)}

# contaminated component
contrIG    <- function(x, t, g, a, e, log = F){ (1-a)*dgamma(x,t/(g*e)+1,g*e,log=log)}

z <- function(x, t, g, a, e){
  return(ifelse((a*rIG(x, t, g))/(clnrIG(x, t, g, a, log = F) + contrIG(x, t, g, a, e, log = F)) > 0.5, 1, 0))
}

v <- 0.4
theta <- 2
alpha <- 0.85
eta <- 4

set.seed(18)
sim <- x

good <- z(sim, 2, 0.4, 0.85, 4)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.85, 2, 0.4, 4)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.85
theta_1 <- 2
gamma_1 <- 0.4
eta_1 <- 4


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0000001 & k <= 150) {
  # E step
  good <- z(sim,theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dmgammamixecm <- function(par){
  t <- exp(par[1])
  g <- exp(par[2])
  e <- exp(par[3])
  sum(good*rIG(sim, t, g, log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- exp(optim(log(c(theta_1, gamma_1, eta_1)), dmgammamixecm, control=list(fnscale= -1))$par[1:3])
   search[k, 6] <- optim(log(c(theta_1, gamma_1, eta_1)), dmgammamixecm, control=list(fnscale= -1))$value
  
  alpha_1 <- search[k,2]
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  loglike <- search[k, 6]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}

hist(sim, main = "", xlim = c(75, 200), freq = FALSE, breaks = 20)
lines(density(sim))
rug(sim,side=1,col="black")
#curve(clnrIG(x, 2, 0.4, a = 0.85) + contrIG(x, 2, 0.4, a = 0.85, 4), 0, 50, ylab = "", ylim = c(0, 0.1))
curve(clnrIG(x, theta_1, gamma_1, a = alpha_1), 80, 200, add = TRUE, lty = 2)
curve(contrIG(x, theta_1, gamma_1, a = alpha_1, eta_1), 80, 200, add = TRUE, lty = 3)
legend("topright", legend =c("Actual data density", "good points density", "bad points density"), lty = c(1,2, 3), title = "Simulated & Estimated rIG") 

```



























# Mixed simulation and estimation

```{r}
# define mode parameterised gamma density function
dmgamma <- function(x,theta,v,log=F){
dgamma(x,theta/v+1,v,log=log)
}


theta <- 2
v <- 0.4
eta <- 2 
alpha <- 0.75

#simulate contaminated data 

set.seed(18)
x<- alpha*rgamma(500,theta/v+1,v) + 0.25*rgamma(500,theta/2*v+1,2*v) # because we choose eta = 2

dmgammaMLE <- function(par,x){
lambda <- exp(par[1]) #back-transformation
v <- exp(par[2]) #back-transformation
return(sum(dmgamma(x,lambda,v,log=T)))
}


# we need to use constrained optimisation.
start <- log(c(theta,v)) #transformation


# we can use optim to get the mle's
# optim needs the arguement of the parameters it is trying to estimate as the first argument.
# we use fnscale=-1, because optim minimizes by default.

est <- optim(par = start,fn = dmgammaMLE,x=x, control=list(fnscale= -1))
#extract estimated parameters
par <- est$par
theta <- exp(par[1]) #back transformation
vest <- exp(par[2])
loglike <- est$value

curve

```

# Contaminated simulation and contaminated estimation 

```{r,  warning = FALSE}

# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}

# reparameterised rIG pdf 
rIG <- function(x, t, g, log = F){dgamma(x,t/g+1,g,log=log)}

crIG <- function(x, t, g, a, e, log = F){
  return(ifelse( a >= 1, dgamma(x,t/g+1,g,log=log), a*dgamma(x,t/g+1,g,log=log) + (1-a)*dgamma(x,theta/(g*e)+1,g*e,log=log)))
}

#clean component
clnrIG    <- function(x, t, g, a , log = F){ (a)*dgamma(x,t/g+1,g,log=log)}

# contaminated component
contrIG    <- function(x, t, g, a , e, log = F){ (1-a)*dgamma(x,t/g*e+1,g*e,log=log)}

z <- function(x, t, g, a, e){
  return(ifelse(crIG(x, t, g, a, e) <= 0, 1, ifelse((a*rIG(x, t, g))/crIG(x, t, g, a, e) > 0.5, 1, 0)))
}

v <- 0.4
theta <- 2
alpha <- 0.85
eta <- 4

set.seed(18)
sim <- c(rgamma(500*0.75,theta/v+1,v), rgamma(500*0.25,theta/(2*v)+1,2*v))


good <- z(sim[1:500], 2, 0.4, 0.85, 4)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.75, 2, 0.4, 2, 0)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.75
theta_1 <- 2
gamma_1 <- 0.4
eta_1 <- 2


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0001) {
  # E step
  good <- z(sim[1:500],theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dmgammamixecm <- function(par){
  t <- par[1]
  g <- par[2]
  e <- par[3]
  sum.finite(good*rIG(sim, t, g,  log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- optim(log(c(theta_1, gamma_1, eta_1)), dmgammamixecm)$par[1:3]
  
  alpha_1 <- search[k,2]
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}

curve(clnrIG(x, 2, 0.4) + contrIG(x, 2, 0.4, a = 0.75, 2), 0, 100)
curve(crIG(x, 2, 0.4), 0, 100)

```



```{r,  warning = FALSE}

# modified sum only considers finite values
sum.finite <- function(x) {
  sum(x[is.finite(x)])
}

# reparameterised rIG pdf 
rIG <- function(x, t, g, log = F){dgamma(x,t/g+1,g,log=log)}

#clean component
clnrIG    <- function(x, t, g, a, log = F){ (a)*dgamma(x,t/g+1,g,log=log)}

# contaminated component
contrIG    <- function(x, t, g, a, e, log = F){ (1-a)*dgamma(x,t/(g*e)+1,g*e,log=log)}

z <- function(x, t, g, a, e){
  return(ifelse((a*rIG(x, t, g))/(clnrIG(x, t, g, a, log = F) + contrIG(x, t, g, a, e, log = F)) > 0.5, 1, 0))
}

v <- 0.4
theta <- 2
alpha <- 0.75
eta <- 2

set.seed(18)
sim <- c(rgamma(500*0.75,theta/v+1,v), rgamma(500*0.25,theta/(2*v)+1,2*v))

good <- z(sim, 2, 0.4, 0.75, 2)

Q <- data.frame(matrix(,nrow = 0, ncol = 2))

Q[1,] <- cbind(1,0)

# starting value of expected value of the log likelihood
Q[2,] <- cbind(2,sum.finite(good*log(alpha) + (1-good)*log(1-alpha)) + sum.finite(good*rIG(sim,theta,v, log = T) + (1-good)*contrIG(sim,theta, v, a = 0, eta, log = T)))

k <- 2

search <- data.frame(matrix(, ncol = 5, nrow = 0))
colnames(search) <- c("Iteration", "alpha", "theta", "gamma", "eta")
search[1,] <- rbind(1, 0.75, 2, 0.4, 2)

# Prepare parameters for implementation of EM Algorithm
alpha_1 <- 0.75
theta_1 <- 2
gamma_1 <- 0.4
eta_1 <- 2


while (abs(Q[k, 2]-Q[k-1, 2])>=0.0000001) {
  # E step
  good <- z(sim,theta_1, gamma_1, alpha_1, eta_1)
  
  # M step
  dmgammamixecm <- function(par){
  t <- exp(par[1])
  g <- exp(par[2])
  e <- exp(par[3])
  sum(good*rIG(sim, t, g, log = T) + (1-good)*contrIG(sim, t, g, a = 0 , e, log = T))}
  
  
  # Update iteration tracker
  search[k, 1] <- k 
  
  # Update alpha
  search[k,2] <- max(0.5, sum(good == 1)/length(sim)) # mean(z_i)
  
  # Update theta, gamma, eta 
  search[k, 3:5] <- exp(optim(log(c(theta_1, gamma_1, eta_1)), dmgammamixecm, control=list(fnscale= -1))$par[1:3])
  
  alpha_1 <- search[k,2]
  theta_1 <- search[k,3]
  gamma_1 <- search[k,4]
  eta_1 <- search[k,5]
  
  k <- k + 1
  Q[k, ] <- cbind(k, sum.finite(good*log(alpha_1) + (1-good)*log(1-alpha_1)) + sum.finite(good*rIG(sim,theta_1,gamma_1,log = T)) + (1-good)*contrIG(sim,theta_1, gamma_1, 0, eta_1, log = T)) 
  
  
}

curve(clnrIG(x, 2, 0.4, a = 0.75) + contrIG(x, 2, 0.4, a = 0.75, 2), 0, 50, ylab = "", ylim = c(0, 0.1))
curve(clnrIG(x, theta_1, gamma_1, a = alpha_1) + contrIG(x, theta_1, gamma_1, a = alpha_1, eta_1), 0, 50, add = TRUE, lty = 2)

```